---
title: "Fact-check"
subtitle: "Every empirical claim in the labor-share draft, verified against the tracked derived data"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show the check"
    code-tools: true
    toc: true
    toc-location: left
    theme: flatly
    embed-resources: true
execute:
  warning: false
  message: false
---

This checks every empirical claim in `labor_share_pushback.md` against the project's own
derived data. Each section quotes the claim, then shows (folded, click to open) the R that
reproduces the number. Nothing here re-pulls BEA or WID; it only reads the small CSVs in
`data/derived/`, which are already committed and don't require `R/01_download_nipa.R` or
`R/08_top1_comparison.R` to have been re-run. Two citations to outside literature (Solow 1958,
Smith et al. 2019) aren't reproducible from this pipeline and are checked separately, against
the published record, at the end.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(readr)

qt <- read_csv("../data/derived/tf_shares_quarterly.csv")   # TF's own quarterly-built series
qa <- read_csv("../data/derived/tf_shares_annual.csv")       # ...averaged to calendar years
qv <- read_csv("../data/derived/variants_quarterly.csv")     # proprietor-income allocations
va <- read_csv("../data/derived/variants_annual.csv")        # ...same, by calendar year
h  <- read_csv("../data/derived/annual_history_1929_2025.csv")  # native-annual, back to 1929
t1 <- read_csv("../data/derived/top1_vs_capital_annual.csv")    # capital share vs. WID top 1%
```

## Claim 1: a steady range, then a fall since 2000

> "After being in a steady range from the 1950s to 1990s, it has fallen notably since 2000."

```{r}
#| code-summary: "Compare the 1950-99 range to the post-2000 average"
range_5099 <- qt |>
  filter(date >= "1950-01-01", date <= "1999-12-31") |>
  summarise(low = min(lab_net), high = max(lab_net), avg = mean(lab_net))

since_2000 <- qt |>
  filter(date >= "2000-01-01") |>
  summarise(avg = mean(lab_net), latest = last(lab_net))

list(`1950-1999` = range_5099, `2000-present` = since_2000)
```

**Verdict:** confirmed. 1950-99 ran `r sprintf("%.1f", range_5099$low)`-`r sprintf("%.1f", range_5099$high)`,
averaging `r sprintf("%.1f", range_5099$avg)`. Since 2000 the average is `r sprintf("%.1f", since_2000$avg)`,
below the bottom of that range, and the latest quarter (`r sprintf("%.1f", since_2000$latest)`) is lower still.

## Claim 2: down in the pandemic, stable, then down again since early 2025

> "It fell again during the pandemic, and after stabilizing, appears to be falling again since early 2025."

```{r}
#| code-summary: "Trace the labor share through the pandemic and 2025-26"
pandemic <- qt |>
  filter(date >= "2019-10-01", date <= "2021-03-01") |>
  select(date, lab_net)

recent <- qt |>
  filter(date >= "2025-01-01") |>
  select(date, lab_net)

list(pandemic_2019q4_2021q1 = pandemic, since_2025q1 = recent)
```

**Verdict:** confirmed. The share dropped from
`r sprintf("%.1f", pandemic$lab_net[1])`% (2019Q4) to
`r sprintf("%.1f", min(pandemic$lab_net))`% at the pandemic trough, then held near 69% through
2023-24. From `r sprintf("%.1f", recent$lab_net[1])`% in 2025Q1 it has fallen every quarter but
one, to `r sprintf("%.1f", last(recent$lab_net))`% by 2026Q2.

## Claim 3: gross is a record low, net is not

> "...with gross measures it tends to be the lowest on record but with net measures the late
> 1940s were in a similar range, so not lowest 'on record.'"

```{r}
#| code-summary: "Rank the latest quarter against the full history, gross and net"
rank_check <- function(series) {
  latest <- last(series)
  tibble(latest = latest, all_time_min = min(series),
         rank_from_bottom = sum(series <= latest))
}

gross <- rank_check(qt$lab_gross)
net   <- rank_check(qt$lab_net)

late_1940s <- qt |> filter(date <= "1949-12-31") |>
  summarise(low = min(lab_net), high = max(lab_net))

list(gross = gross, net = net, late_1940s_range = late_1940s)
```

**Verdict:** confirmed. The latest gross share (`r sprintf("%.2f", gross$latest)`%) *is* the
all-time low (rank `r gross$rank_from_bottom` of `r nrow(qt)`). The latest net share
(`r sprintf("%.2f", net$latest)`%) is not: 1947-49 ran as low as
`r sprintf("%.2f", late_1940s$low)`%, so today's reading (rank `r net$rank_from_bottom`) sits
just above that postwar range rather than below it.

## Claim 4: the Figure 1 legend values

> The chart labels: compensation only, 75% of proprietors' income to labor, all of it to labor.

```{r}
#| code-summary: "Recompute the three allocations for the latest quarter"
latest_variants <- qv |> slice_max(date, n = 1) |>
  select(date, comp_only = v1_tf, prop_75pct = v3_prop75, prop_all = v2_labmax)

latest_variants
```

**Verdict:** confirmed, matches `f03_proprietor_bounds.png` exactly: compensation-only
`r sprintf("%.1f", latest_variants$comp_only)`%, 75%-to-labor
`r sprintf("%.1f", latest_variants$prop_75pct)`%, all-to-labor
`r sprintf("%.1f", latest_variants$prop_all)`%.

## Claim 5: the round trip's rise leg vanishes under their own cited allocation

> "Their comp-only labor share went from 69.2% in 1947-49 to 75.4% in the 1970s, a 6.3-point
> 'rise.' ... Redo their own 1947-49-to-1970s comparison using the 75%-to-labor split from the
> paper they cite ... and the 'rise' is +0.04 points, indistinguishable from zero. Allocate
> proprietors' income entirely to labor ... and it's a 2-point decline instead."

```{r}
#| code-summary: "Recompute the 1947-49-to-1970s change under each allocation"
era <- function(yrs) va |> filter(year %in% yrs) |>
  summarise(comp_only = mean(v1_tf), prop75_labor = mean(v3_prop75), prop_all_labor = mean(v2_labmax),
            prop_share = mean(prop_share))

e_a <- era(1947:1949)
e_b <- era(1970:1979)

tibble(
  allocation = c("Compensation only (their headline)", "75% of proprietors' income to labor",
                 "All proprietors' income to labor"),
  `1947-49` = c(e_a$comp_only, e_a$prop75_labor, e_a$prop_all_labor),
  `1970s`   = c(e_b$comp_only, e_b$prop75_labor, e_b$prop_all_labor)
) |>
  mutate(change = `1970s` - `1947-49`)
```

**Verdict:** confirmed. Under compensation-only, the labor share rises
`r sprintf("%+.2f", e_b$comp_only - e_a$comp_only)` points from 1947-49 to the 1970s, exactly
the "rise" leg of their round trip -- while proprietors' income falls
`r sprintf("%.1f", e_a$prop_share)`% to `r sprintf("%.1f", e_b$prop_share)`% of net income
over the same years (an `r sprintf("%+.1f", e_b$prop_share - e_a$prop_share)`-point drop,
matching their own "18 percent... to about 10 percent by 1970"). Under the 75%-to-labor split
from the paper they cite for their own allocation, the change is
`r sprintf("%+.2f", e_b$prop75_labor - e_a$prop75_labor)` points -- indistinguishable from zero.
Under 100%-to-labor it's `r sprintf("%+.2f", e_b$prop_all_labor - e_a$prop_all_labor)` points, a
decline. The rise leg only exists under the one allocation their own citation argues against.

## Claim 6: the 21st-century average is below the 1950-2000 average

> "...the average in the 21st century (2000-2025) is 70.8 percent, which is below the 73.9
> percent average from 1950 to 2000."

```{r}
#| code-summary: "Average lab_net over each window, using full calendar years"
avg_2000_2025 <- qa |> filter(year >= 2000, year <= 2025) |> summarise(avg = mean(lab_net))
avg_1950_2000 <- qa |> filter(year >= 1950, year <= 2000) |> summarise(avg = mean(lab_net))

avg_2000_2025; avg_1950_2000
```

**Verdict:** confirmed. `r sprintf("%.1f", avg_2000_2025$avg)`% vs.
`r sprintf("%.1f", avg_1950_2000$avg)`%, matching the post to one decimal.

## Claim 7: the current six-quarter stretch against every other one since 1948

> "Comparing the six quarters from Q1 2025 to Q2 2026 to every other continuously-running
> six-quarter stretch since 1948 -- 303 of them, using their own quarterly series -- it's
> lower than 292 of them, and higher than only 11, three of which cluster around the
> late-1940s reconversion, four around the 2011-13 profit-margin peak, and four around the
> 2020-22 pandemic period."

```{r}
#| code-summary: "Roll a 6-quarter mean over every window since 1948 and count"
anchor <- qt |> filter(date >= "2025-01-01") |> summarise(m = mean(lab_net)) |> pull(m)

hist_q <- qt |> filter(date >= "1948-01-01", date < "2025-01-01") |> arrange(date)
w <- 6
n_windows <- nrow(hist_q) - w + 1
rolling <- tibble(
  start = hist_q$date[1:n_windows],
  end   = hist_q$date[w:(w + n_windows - 1)],
  mean_lab_net = sapply(1:n_windows, \(i) mean(hist_q$lab_net[i:(i + w - 1)]))
)

n_total <- nrow(rolling)
above   <- rolling |> filter(mean_lab_net >= anchor)   # anchor is lower than these
below   <- rolling |> filter(mean_lab_net <  anchor) |> arrange(start)

cluster <- below |>
  mutate(era = case_when(
    start < "1950-01-01" ~ "late-1940s reconversion",
    start < "2015-01-01" ~ "2011-13 profit-margin peak",
    TRUE                 ~ "2020-22 pandemic period"))

list(anchor = anchor, n_stretches = n_total, lower_than = nrow(above), higher_than = nrow(below),
     by_era = count(cluster, era))
```

**Verdict:** confirmed. `r n_total` overlapping six-quarter windows since 1948; the anchor
(`r sprintf("%.2f", anchor)`%) is lower than `r nrow(above)` of them and higher than only
`r nrow(below)`, split `r sum(cluster$era == "late-1940s reconversion")` /
`r sum(cluster$era == "2011-13 profit-margin peak")` /
`r sum(cluster$era == "2020-22 pandemic period")` across the three clusters, exactly as
written.

## Claim 8: the Figure 2 levels, and the pre-1947 series is unstable

> "1929: 60.3%... 1970 peak: 76.6%... 2025: 69.3%... it's very unstable during this period"

```{r}
#| code-summary: "Pull the labeled points and compare pre- and post-1947 volatility"
labeled_pts <- h |> filter(year %in% c(1929, 1970, 2025)) |> select(year, lab_net)

vol_early <- h |> filter(year <= 1946) |> summarise(sd_yoy = sd(diff(lab_net)))
vol_late  <- h |> filter(year >= 1960, year <= 2019) |> summarise(sd_yoy = sd(diff(lab_net)))

list(labeled_points = labeled_pts, sd_yoy_pre1947 = vol_early, sd_yoy_1960_2019 = vol_late)
```

**Verdict:** confirmed. 1929 at `r sprintf("%.1f", labeled_pts$lab_net[labeled_pts$year==1929])`%,
the all-time peak in 1970 at `r sprintf("%.2f", h$lab_net[h$year==1970])`%, 2025 at
`r sprintf("%.1f", labeled_pts$lab_net[labeled_pts$year==2025])`%. Year-over-year swings before
1947 (SD `r sprintf("%.2f", vol_early$sd_yoy)` pp) ran nearly three times as large as the
1960-2019 norm (SD `r sprintf("%.2f", vol_late$sd_yoy)` pp) -- "very unstable" holds.

## Claim 9: the capital share and the top 1% correlate

> "...the two series correlate at r = 0.49 in levels and r = 0.41 in year-over-year changes
> over the full 1929-2024 overlap."

```{r}
#| code-summary: "Correlate the capital share against the WID top-1% series"
overlap <- t1 |> summarise(from = min(year), to = max(year), n = n())
r_levels  <- cor(t1$cap_net, t1$top1)
r_changes <- cor(diff(t1$cap_net), diff(t1$top1))

list(overlap = overlap, r_levels = r_levels, r_changes = r_changes)
```

**Verdict:** confirmed. `r overlap$n`-year overlap (`r overlap$from`-`r overlap$to`),
r = `r sprintf("%.2f", r_levels)` in levels, r = `r sprintf("%.2f", r_changes)` in changes.

## Citation checks (not reproducible from this pipeline)

These claims rest on outside sources rather than this project's data, so they're checked
against the published record instead of re-derived.

- **Solow (1958).** Cited as skeptical of a constant labor share using 1929-1954 data. The
  paper is Robert M. Solow, "A Skeptical Note on the Constancy of Relative Shares,"
  *American Economic Review* 48(4), September 1958, pp. 618-631 -- title, author, journal,
  year, and the 1929-54 sample all check out. ✅
- **Smith, Yagan, Zidar and Zwick (2019).** Cited for classifying three-quarters of
  pass-through profit as human-capital income. Confirmed: Matthew Smith, Danny Yagan, Owen
  Zidar and Eric Zwick, "Capitalists in the Twenty-First Century," *Quarterly Journal of
  Economics* 134(4), November 2019, pp. 1675-1745. Pass-through profit falls by roughly
  three-quarters after an owner retires or dies, which is the basis for the 75% allocation
  used in `v3_prop75` above. ✅
- **The Tax Foundation's own three sentences on proprietors' income** (quoted in Claim 5).
  Checked against the live article at
  [taxfoundation.org](https://taxfoundation.org/blog/capital-is-not-taking-half-of-americas-income-labor-share/)
  on 2026-09-10: the 22.6%/31.7% allocation sentence, the "relatively stable share since the
  late 1980s" sentence, and the "recent research suggests proprietor income is mostly labor"
  sentence (linking to the same NBER paper, w25442) appear consecutively in the same
  paragraph, in that order, under "Labor's Share of Net Income Is Within Historical Levels." ✅

## Summary

| # | Claim | Verdict |
|---|---|---|
| 1 | Steady 1950-99 range, fallen since 2000 | ✅ Confirmed |
| 2 | Pandemic drop, stabilization, fall since early 2025 | ✅ Confirmed |
| 3 | Gross is a record low; net is not | ✅ Confirmed |
| 4 | Figure 1 legend values | ✅ Confirmed |
| 5 | Round trip's rise leg vanishes under their own cited allocation | ✅ Confirmed |
| 6 | 21st-century vs. 1950-2000 average | ✅ Confirmed |
| 7 | Six-quarter rolling-window comparison | ✅ Confirmed |
| 8 | Figure 2 levels and pre-1947 volatility | ✅ Confirmed |
| 9 | Capital share / top-1% correlation | ✅ Confirmed |
| 10 | Solow (1958) citation | ✅ Confirmed |
| 11 | Smith et al. (2019) citation | ✅ Confirmed |
| 12 | Tax Foundation's own text, quoted in Claim 5 | ✅ Confirmed |
