---
title: "How you measure it: a survey of labor-share and corporate-profit conventions"
subtitle: "Quarterly NIPA (and BLS) data only, 1947Q1 forward -- no annual aggregation"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    code-tools: true
    toc: true
    toc-location: left
    theme: flatly
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(scales)
library(gt)

panel <- read_csv("../data/derived/measures_survey_quarterly.csv", show_col_types = FALSE)
meta  <- read_csv("../data/derived/measures_survey_meta.csv",      show_col_types = FALSE)
rk    <- read_csv("../data/derived/measures_survey_ranking.csv",   show_col_types = FALSE)

last_date <- max(panel$date)
last_q    <- paste0(format(last_date, "%Y"), "Q", (as.integer(format(last_date, "%m")) + 2) %/% 3)
first_q   <- paste0(format(min(panel$date), "%Y"), "Q1")

theme_ls <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(linewidth = 0.25, colour = "grey85"),
          plot.title    = element_text(face = "bold", size = rel(1.05)),
          plot.subtitle = element_text(colour = "grey30", size = rel(0.85)),
          plot.caption  = element_text(colour = "grey45", hjust = 0, size = rel(0.72)))
}
wrp <- function(x, n = 100) paste(strwrap(x, width = n), collapse = "\n")

plot_measure <- function(id_, extreme = c("low", "high")) {
  extreme <- match.arg(extreme)
  x  <- panel |> filter(id == id_) |> arrange(date)
  mr <- meta  |> filter(id == id_)
  r  <- rk    |> filter(id == id_)
  is_index <- mr$unit == "Index, 2017=100"
  col <- if (mr$group == "Labor share") "#1B4F72" else "#B54708"

  ref_date  <- if (extreme == "low") r$record_date else r$record_high_date
  ref_value <- if (extreme == "low") r$record_value else r$record_high_value
  ref_lab   <- if (extreme == "low") "record low" else "record high"

  p <- ggplot(x, aes(date, value)) +
    geom_hline(yintercept = ref_value, linetype = "22", colour = "grey55") +
    geom_line(linewidth = 0.7, colour = col) +
    geom_point(data = x |> slice_max(date, n = 1), size = 1.8, colour = col) +
    annotate("text", x = ref_date, y = ref_value,
             label = sprintf("%s: %.1f (%s)", ref_lab, ref_value,
                              paste0(format(ref_date, "%Y"), "Q", (as.integer(format(ref_date, "%m")) + 2) %/% 3)),
             hjust = 0.05, vjust = -0.6, size = 2.9, colour = "grey35") +
    scale_x_date(breaks = seq(as.Date("1950-01-01"), last_date, by = "10 years"), date_labels = "%Y") +
    labs(title = NULL,
         subtitle = sprintf("%s | %s | latest (%s): %.1f", mr$sector, mr$basis, last_q, r$latest_value),
         x = NULL, y = mr$unit,
         caption = paste0("Source: ", mr$source, ". Quarterly, ", first_q, "-", last_q, ".")) +
    theme_ls()

  if (is_index) {
    p <- p + scale_y_continuous(labels = label_number(accuracy = 1))
  } else {
    p <- p + scale_y_continuous(labels = label_percent(scale = 1))
  }
  p
}

make_table <- function(ids, extreme = c("low", "high")) {
  extreme <- match.arg(extreme)
  d <- rk |> filter(id %in% ids) |> left_join(meta |> select(id, description), by = "id")
  d <- d[match(ids, d$id), ]
  if (extreme == "low") {
    d <- d |> transmute(Name = name, Description = description,
                         `Is it the lowest?` = lowest_label, Context = context)
  } else {
    d <- d |> transmute(Name = name, Description = description,
                         `Is it the highest?` = highest_label, Context = context_high)
  }
  d |> gt() |>
    tab_options(table.font.size = 13, data_row.padding = px(6)) |>
    cols_width(Name ~ px(170), Description ~ px(340),
               starts_with("Is it") ~ px(120), Context ~ px(340)) |>
    opt_row_striping()
}
```

Every "labor share is falling" or "labor share made a round trip" argument rests on a choice
of numerator, denominator and universe that the data itself doesn't dictate. This note lays
out the choices side by side rather than picking one. It's built to sit next to
`docs/replication_and_literature.md`, which replicates and stress-tests the Tax Foundation's
own specific convention; this note instead surveys the wider menu, including the four
different conventions charted (without being reconciled) in an internal deck,
`draft_deck_labor_share_trend_mkonczal (2).pptx`.

**Scope.** Quarterly only, `r first_q`-`r last_q` -- NIPA Table 1.10 and 1.14 are published
quarterly back to 1947Q1, so that's the full available window; nothing here is annualized.
Two axes run through every measure below:

- **Gross vs. net.** Whether depreciation (consumption of fixed capital) is removed from the
  income base. Net is the standard correction in the literature
  ([Rognlie 2015](https://www.brookings.edu/wp-content/uploads/2016/07/2015a_rognlie.pdf);
  [Bridgman 2018](https://www.cambridge.org/core/journals/macroeconomic-dynamics/article/abs/is-labors-loss-capitals-gain-gross-versus-net-labor-shares/957C5C3D90762547B3CF4924EF34F4E9)),
  and changes the *level* a lot; whether it changes the *trend* is measure-specific and shown
  below for each one.
- **Universe.** Economy-wide (NIPA Table 1.10), the BLS's own official nonfarm-business
  statistic, all corporations (NIPA Table 1.14), or nonfinancial corporations only (also
  Table 1.14, and the internal deck's preferred universe since it has no proprietors' income
  or imputed rent to allocate).

Within the economy-wide universe, proprietors' income is genuinely ambiguous -- it's a mix of
labor and capital return that NIPA doesn't split
([Gollin 2002](https://www.journals.uchicago.edu/doi/10.1086/338747)) -- so three allocations
are shown: all of it counted as capital (the narrowest labor share), 75% to labor
([Smith, Yagan, Zidar and Zwick 2019](https://academic.oup.com/qje/article-abstract/134/4/1675/5542244)),
and all of it counted as labor (the widest). Corporate-sector measures don't have this
problem: wages on one line, profits on another.

Every number below is a direct BEA NIPA or BLS series or ratio of them -- nothing is
estimated, smoothed or seasonally-readjusted beyond what BEA/BLS already publish. Full
definitions and series codes are in `R/10_measures_survey.R`.

## Labor share

### Economy-wide

```{r}
#| results: asis
for (id_ in c("l_gross_econ", "l_net_econ_std", "l_net_tf", "l_net_labmax", "l_net_prop75")) {
  nm <- meta$name[meta$id == id_]
  cat(sprintf("\n\n#### %s\n\n", nm))
  print(plot_measure(id_, "low"))
}
```

### BLS official measure

```{r}
#| results: asis
cat(sprintf("\n\n#### %s\n\n", meta$name[meta$id == "l_bls_nfb"]))
print(plot_measure("l_bls_nfb", "low"))
```

Note the units: this is BLS's own index (2017 = 100), not a percent of income. Ranking it
against its own history is still meaningful -- it's a fixed-base rescaling of the same
compensation-to-output ratio -- but the level isn't comparable to the percent-of-income
measures above or below.

### Corporate business (all corporations)

```{r}
#| results: asis
for (id_ in c("l_corp_gross", "l_corp_net_std", "l_corp_net_extax")) {
  cat(sprintf("\n\n#### %s\n\n", meta$name[meta$id == id_]))
  print(plot_measure(id_, "low"))
}
```

### Nonfinancial corporate business

The internal deck's preferred universe (slides 5 and 16): no proprietors' income, no
financial-sector profits, no imputed housing rent to argue about.

```{r}
#| results: asis
for (id_ in c("l_nfcorp_gross", "l_nfcorp_net_std", "l_nfcorp_net_extax")) {
  cat(sprintf("\n\n#### %s\n\n", meta$name[meta$id == id_]))
  print(plot_measure(id_, "low"))
}
```

### Labor share: is it a record low?

```{r}
labor_order <- c("l_gross_econ", "l_net_econ_std", "l_net_tf", "l_net_labmax", "l_net_prop75",
                  "l_bls_nfb", "l_corp_gross", "l_corp_net_std", "l_corp_net_extax",
                  "l_nfcorp_gross", "l_nfcorp_net_std", "l_nfcorp_net_extax")
make_table(labor_order, "low")
```

```{r}
oxford_and <- function(x) {
  if (length(x) <= 1) return(paste(x, collapse = ""))
  paste0(paste(x[-length(x)], collapse = ", "), " and ", x[length(x)])
}
lab_rk        <- rk |> filter(id %in% labor_order)
n_labor_low   <- sum(lab_rk$is_record_low)
not_low_names <- lab_rk$name[!lab_rk$is_record_low]
```

`r n_labor_low` of the `r length(labor_order)` labor-share measures above are sitting at
their record low of the `r first_q`-`r last_q` quarterly series. The `r length(labor_order) - n_labor_low`
that aren't -- `r oxford_and(not_low_names)` -- are each within about two points of their own
record, not comfortably inside their historical range.

## Corporate profits

The mirror question: a falling labor share and a rising profit share are largely the same
fact viewed from opposite sides of the same ratio (compensation + capital income + taxes sum
to the whole by construction), so the tables below ask about **record highs**, not record
lows.

### Economy-wide

```{r}
#| results: asis
cat(sprintf("\n\n#### %s\n\n", meta$name[meta$id == "p_all_pretax_gdi"]))
print(plot_measure("p_all_pretax_gdi", "high"))
```

### Corporate business (all corporations)

```{r}
#| results: asis
for (id_ in c("p_all_pretax_gross", "p_all_pretax_net", "p_all_aftertax_gross", "p_all_aftertax_net")) {
  cat(sprintf("\n\n#### %s\n\n", meta$name[meta$id == id_]))
  print(plot_measure(id_, "high"))
}
```

### Nonfinancial corporate business

Slide 2 of the internal deck charts the first of these (pretax, gross) to argue profits
"remain elevated" post-pandemic; the other three vary the tax and depreciation treatment.

```{r}
#| results: asis
for (id_ in c("p_nf_pretax_gross", "p_nf_pretax_net", "p_nf_aftertax_gross", "p_nf_nos_net")) {
  cat(sprintf("\n\n#### %s\n\n", meta$name[meta$id == id_]))
  print(plot_measure(id_, "high"))
}
```

### Corporate profits: is it a record high?

```{r}
profit_order <- c("p_all_pretax_gdi", "p_all_pretax_gross", "p_all_pretax_net",
                   "p_all_aftertax_gross", "p_all_aftertax_net",
                   "p_nf_pretax_gross", "p_nf_pretax_net", "p_nf_aftertax_gross", "p_nf_nos_net")
make_table(profit_order, "high")
```

```{r}
prof_rk        <- rk |> filter(id %in% profit_order)
n_profit_high  <- sum(prof_rk$is_record_high)
high_names     <- prof_rk$name[prof_rk$is_record_high]
nf_rk          <- prof_rk |> filter(sector == "Nonfinancial corporate")
nf_rank_range  <- range(nf_rk$rank_high)
```

`r n_profit_high` of the `r length(profit_order)` profit measures -- `r oxford_and(high_names)`
-- are at their record high. The nonfinancial-only cuts, which is the universe the internal
deck actually uses, are elevated (`r nf_rank_range[1]`nd-`r nf_rank_range[2]`th highest of
`r prof_rk$n[1]` quarters) but have each been marginally exceeded at some point since 1947, mostly around the
2011-13 and 2021-22 profit-margin peaks.

## What this does and doesn't settle

Twelve labor-share conventions and nine corporate-profit conventions, spanning gross/net,
four sector universes and three proprietor-income allocations, agree that both series sit at
or extremely near an 80-year extreme as of `r last_q`. That's a statement about levels. It
says nothing about whether the capital income these measures net out is a competitive return
or economic rent -- see `docs/replication_and_literature.md` for the markup and superstar-firm
literature that speaks to that question, which an aggregate income-share identity can't
answer on its own.
